`r params$doc_title`

### 2022-12-16 scripts/MSc.Rmd
added ambiguous FTO variant to analysis

my_data_harm <- my_data_harm %>% 
  filter(mr_keep==TRUE)
```

@@

## CALCULATE explained variance
my_data_harm <- my_data_harm %>% 
  mutate(explained_variance_BMI = explained_variance_numeric(eaf = eaf.bmi, beta = beta.bmi), 
         explained_variance_BMI2 = explained_variance_numeric2(maf=eaf.bmi, 
                                                           beta=beta.bmi, 
                                                           se_beta=se.bmi, 
                                                           samplesize=n[length(n)]

         ),
         explained_variance_GSD = explained_variance_binary(PA = eaf.gsd, 
                                                            RR1 = exp(beta.gsd), 
                                                            RR2 = exp(beta.gsd)^2, 
                                                            K = 0.1  # assumed prevalence in population
                                                            )$Vg,
         explained_variance_GBC2 = explained_variance_binary(PA = eaf.gbc, 
                                                            RR1 = exp(beta.gbc), 
                                                            RR2 = exp(beta.gbc)^2, 
                                                            K =1.2/100000  # assumed prevalence in population
                                                            )$Vg,
         explained_variance_GBC = explained_variance_binary(PA = eaf.gbc, 
                                                            RR1 = exp(beta.gbc), 
                                                            RR2 = 1+(2*(exp(beta.gbc)-1)), 
                                                            K =1.2/100000  # assumed prevalence in population
                                                            )$Vg

         )

sum(my_data_harm$explained_variance_GBC)
sum(my_data_harm$explained_variance_GBC2)
```

@@

```{r PVE_BMI_vs_pval, warning=FALSE}
# labeling (only FTO label)
my_data_harm$gene_label <- NA
my_data_harm$gene_label[my_data_harm$SNP %in% filter_FTO$refsnp_id] <- "FTO/MC4R"

@@

for (i in 1:length(colnames_pval) ) {
  threshold <- 5*10^(-6)
  select_col <- colnames_pval[i]
  # calculate cum explained variance and count of IVs
  summarydat <- my_data_harm %>% 
    filter(!!sym(select_col) < threshold) %>% 
    select(explained_variance_BMI, explained_variance_BMI2, explained_variance_GSD, explained_variance_GBC) %>% 
    summarise(expl_variances =across(.cols = everything(), function(x) sum(x, na.rm = TRUE)),
              count_IVs = n())
  # fill up df
  df_cum_expl_var[i, 2] <- summarydat$expl_variances[1]
  df_cum_expl_var[i, 3] <- summarydat$expl_variances[2]
  df_cum_expl_var[i, 4] <- summarydat$expl_variances[3]
  df_cum_expl_var[i, 5] <- summarydat$expl_variances[4]
  df_cum_expl_var[i, 6] <- summarydat$count_IVs
}
```

```{r plot-fig1, warning=FALSE}
# 6.       Represent figure similar to Fig1 in Chatterjee’s paper

# Dual Y axis: we have 284 IVs, scale them to 5% (y-axis [0;5%]): 3/284 
## https://r-graph-gallery.com/line-chart-dual-Y-axis-ggplot2.html
scale_factor <- 4.5/284
df_cum_expl_var <- df_cum_expl_var %>% 
  mutate(
    cum_expl_varX = cum_expl_varX *100, # in percent 
    cum_expl_varX2 = cum_expl_varX2 *100,
    cum_expl_varU = cum_expl_varU *100,
    cum_expl_varY = cum_expl_varY *100,
    count_IVs_scaled = count_IVs*scale_factor)
```

```{r updated-fig01, fig.height=6, fig.width=10, warning=FALSE}
plot_data1 <- df_cum_expl_var %>% 
  pivot_longer(cols = c(cum_expl_varX, cum_expl_varU, cum_expl_varY)) %>% 
  mutate(name = factor(name, labels = c("% var(U)", "% var(X)", "% var(Y)")),
         name = factor(name, levels = c("% var(X)", "% var(Y)", "% var(U)") )
         )

@@

plot_data1b <- df_cum_expl_var %>% 
  pivot_longer(cols = c(cum_expl_varX2, cum_expl_varU, cum_expl_varY)) %>% 
  mutate(name = factor(name, labels = c("% var(U)", "% var(X)", "% var(Y)")),
         name = factor(name, levels = c("% var(X)", "% var(Y)", "% var(U)") )
         )
## plot
ggplot(plot_data1b, aes(x=n, y=value, col=name))+
  geom_point(alpha=0.6)+
  geom_line(alpha=0.6)+
  geom_line(aes(x=n, y=count_IVs_scaled, col="Number of IVs"), linetype="dotdash")+
  scale_x_continuous(name = paste0("Size of sample ", as.roman(1)),  # 2-sample-MR or label "Study Size"
                     labels = comma, 
                     breaks = N_annotation, #c( 10000, n[7:length(n)]),
                     limits = c(10000, 300000))+
  #ylab("% variance explained by IVs")+
  labs(title = "Figure 1",
       col='', 
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  scale_y_continuous(
    # Features of the first axis
    name = "% variance explained by IVs",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./scale_factor, name="Number of IVs"),
    limits = c(0,5)
  ) + 
  theme_few()+
  #scale_colour_few() +
  #scale_color_manual(name = "", values = c("% var(U)" = "red", 
  #                                         "% var(X)" = "blue", 
  #                                         "% var(Y)" = "darkblue", 
  #                                         "Number of IVs" = "black"
  #                                         ))
  #
  scale_color_manual(name = "", values = c("% var(U)" = ggthemes_data$few$colors$Dark[2,2][[1]], 
                                           "% var(X)" = ggthemes_data$few$colors$Dark[3,2][[1]], 
                                           "% var(Y)" = ggthemes_data$few$colors$Dark[4,2][[1]], 
                                           "Number of IVs" = "black"),
                     # order
                     breaks = c( "Number of IVs", "% var(X)", "% var(Y)", "% var(U)")
  ) -> fig01b.update
fig01b.update

@@

>>>>

my_data_harm <- my_data_harm %>% 
  filter(mr_keep==TRUE | SNP == "rs1558902")  # also include FTO SNP
```

@@

## CALCULATE explained variance
my_data_harm <- my_data_harm %>% 
  mutate(explained_variance_BMI = explained_variance_numeric(eaf = eaf.bmi, beta = beta.bmi), 
         # USE THIS !
         explained_variance_BMI2 = explained_variance_numeric2(maf=eaf.bmi, 
                                                           beta=beta.bmi, 
                                                           se_beta=se.bmi, 
                                                           samplesize=n[length(n)]

         ),
         explained_variance_GSD = explained_variance_binary(PA = eaf.gsd, 
                                                            RR1 = exp(beta.gsd), 
                                                            RR2 = 1+(2*(exp(beta.gsd)-1)), 
                                                            K = 0.1  # assumed prevalence in population
                                                            )$Vg,
         explained_variance_GBC = explained_variance_binary(PA = eaf.gbc, 
                                                            RR1 = exp(beta.gbc), 
                                                            RR2 = 1+(2*(exp(beta.gbc)-1)), # exp(beta.gbc)^2, 
                                                            K =1.2/100000  # assumed prevalence in population
                                                            )$Vg

         )
# sum(my_data_harm$explained_variance_GBC)
```

@@

```{r PVE_BMI_vs_pval, warning=FALSE}
# labeling (only FTO label)
my_data_harm$gene_label <- NA
my_data_harm$gene_label[my_data_harm$SNP %in% filter_FTO$refsnp_id] <- "FTO/MC4R"
my_data_harm$gene_label[my_data_harm$SNP == "rs1558902"] <- "FTO"
my_data_harm$gene_label[my_data_harm$SNP == "rs6567160"] <- "MC4R"

@@

for (i in 1:length(colnames_pval) ) {
  threshold <- 5*10^(-6)
  select_col <- colnames_pval[i]
  # calculate cum explained variance and count of IVs
  summarydat <- my_data_harm %>% 
    filter(!!sym(select_col) < threshold) %>% 
    select(explained_variance_BMI, explained_variance_BMI2, explained_variance_GSD, explained_variance_GBC) %>% 
    # cumulative PVE
    summarise(expl_variances =across(.cols = everything(), function(x) sum(x, na.rm = TRUE)),
              count_IVs = n())
  # fill up df
  df_cum_expl_var[i, 2] <- summarydat$expl_variances[1]
  df_cum_expl_var[i, 3] <- summarydat$expl_variances[2]
  df_cum_expl_var[i, 4] <- summarydat$expl_variances[3]
  df_cum_expl_var[i, 5] <- summarydat$expl_variances[4]
  df_cum_expl_var[i, 6] <- summarydat$count_IVs
}
```

```{r updated-fig01, fig.height=6, fig.width=10, warning=FALSE}
# 6.       Represent figure similar to Fig1 in Chatterjee’s paper

# Dual Y axis: we have 284 IVs, scale them to 5% (y-axis [0;5%]): 3/284 
## https://r-graph-gallery.com/line-chart-dual-Y-axis-ggplot2.html
scale_factor <- 4.5/284

df_cum_expl_var <- df_cum_expl_var %>% 
  mutate(
    cum_expl_varX = cum_expl_varX *100, # in percent 
    cum_expl_varX2 = cum_expl_varX2 *100,
    cum_expl_varU = cum_expl_varU *100,
    cum_expl_varY = cum_expl_varY *100,
    count_IVs_scaled = count_IVs*scale_factor)

@@

plot_data1b <- df_cum_expl_var %>% 
  pivot_longer(cols = c(cum_expl_varX2, cum_expl_varU, cum_expl_varY)) %>% 
  mutate(name = factor(name, labels = c("% var(U)", "% var(X)", "% var(Y)")),
         name = factor(name, levels = c("% var(X)", "% var(Y)", "% var(U)") )
         )
## plot
ggplot(plot_data1b, aes(x=n, y=value, col=name))+
  geom_point(alpha=0.6)+
  geom_line(alpha=0.6)+
  geom_line(aes(x=n, y=count_IVs_scaled, col="Number of IVs"), linetype="dotdash")+
  scale_x_continuous(name = paste0("Size of sample ", as.roman(1)),  # 2-sample-MR or label "Study Size"
                     labels = comma, 
                     breaks = N_annotation, #c( 10000, n[7:length(n)]),
                     limits = c(10000, 300000))+
  #ylab("% variance explained by IVs")+
  labs(title = "Figure 1",
       col='', 
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  scale_y_continuous(
    # Features of the first axis
    name = "% variance explained by IVs",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./scale_factor, name="Number of IVs"),
    limits = c(0,5)
  ) + 
  theme_few()+
  #scale_colour_few() +
  #scale_color_manual(name = "", values = c("% var(U)" = "red", 
  #                                         "% var(X)" = "blue", 
  #                                         "% var(Y)" = "darkblue", 
  #                                         "Number of IVs" = "black"
  #                                         ))
  #
  scale_color_manual(name = "", values = c("% var(U)" = ggthemes_data$few$colors$Dark[2,2][[1]], 
                                           "% var(X)" = ggthemes_data$few$colors$Dark[3,2][[1]], 
                                           "% var(Y)" = ggthemes_data$few$colors$Dark[4,2][[1]], 
                                           "Number of IVs" = "black"),
                     # order
                     breaks = c( "Number of IVs", "% var(X)", "% var(Y)", "% var(U)")
  ) -> fig01b.update
fig01b.update  # USE THIS

@@

filename_pveplot_sim.raincloud <- paste0("../output/plots/", Sys.Date(), "_figure_simplot_pve_raincloud.png")
ggsave(filename = filename_pveplot_sim.raincloud, plot = plot_pve_simulated.raincloud, 
       width = 10, height = 6, 
       units = "in"  # default
       )
```
